---
title: "Objects Launched into Space"
subtitle: "INFO 526 - Summer 2024 - Final Project"
author: 
  - name: "Derek Rice"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project description"
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
---

```{r}
#| echo = FALSE
  
if (!require("pacman"))
  install.packages("pacman")

pacman::p_load(tidyverse, 
               dplyr, 
               janitor, 
               lubridate,
               ggplot2,
               readr,
               readxl,
               patchwork,
               gganimate,
               ggrepel,
               spData,
               plotly,
               rnaturalearth
              )


# cite: https://github.com/rfordatascience/tidytuesday/blob/master/data/2024/2024-04-23/readme.md

outer_space_objects <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-04-23/outer_space_objects.csv')

#view(outer_space_objects)

objects <- outer_space_objects |>
  na.omit() |>
  mutate(hover = paste0(Entity, "\n", num_objects))


view(objects)

```

```{r}
# cite: https://www.youtube.com/watch?v=RrtqBYLf404
# cite: https://plotly-r.com/maps
# cite: https://plotly.com/r/map-configuration/

objects_graph <- plot_geo(objects,
                          locationmode = "world",
                          frame = ~Year) |>
  add_trace(locations = ~Code,
            z = ~num_objects,
            colorscale = 'Electric',
            text = ~hover,
            hoverinfo = 'text') |>
  layout(title = list(
      text = "<b>Number of Objects Launched into Space by country \n (1957 and 2022)",
      font = list(family = 'Arial', size = 16, color = 'black')   
      ),
           geo = list(
           scope = 'world',
           showland = TRUE,
           landcolor = toRGB("LightGreen"),
           showocean = TRUE,
           oceancolor = toRGB("LightBlue"),
           showcountries = TRUE,
           projection = list(type = "Mercator")),
         font = list(family = 'Arial', size = 12, color = 'black'),
         annotations = list(
           text = "<b>Data Source: tidytuesday,outer_space_objects.csv",
           showarrow = FALSE,
           x = 1,
           y = 0
         )) |>
   config(displayModeBar = FALSE) |>
   colorbar(title = "<b>Number of Objects", color = 'black') 


objects_graph

```


## Abstract



## Introduction

### The Dataset



### Question 1: What nation is likely to dominate space in the not so distant future?

Question 1 - what country is most likely to become the first "space faring nation" and is there a close second? I propose to define "space faring nation" as the country that launches an order of magnitude more vehicles into space than all other countries combined. I'll fit a polynomial regression model to the dataset by country and extrapolate the model into the future to help identify the likely candidates.

## Introduction



## Approach

## Analysis

## Discussion
