---
title: "Objects Launched into Space"
subtitle: "Proposal"
author:
  - name: "Derek Rice"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "INFO526 Term Project"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
if (!require("pacman"))
    install.packages("pacman")

pacman::p_load(tidyverse, tidytuesdayR, dplyr, janitor, lubridate,ggplot2)

```

## Dataset: Outer Space Objects

-   The dataset summarizes the number of objects launched into space from 1957 to 2023 as a function of entity.
-   The source of the dataset is the United Nations Office of Outer Space Affairs. A prior analysis of the dataset is available here: https://ourworldindata.org/grapher/yearly-number-of-objects-launched-into-outer-space
-   Dataset source: https://github.com/rfordatascience/tidytuesday/blob/master/data/2024/2024-04-23/readme.md
-   The dataset is comprised of 4 columns and 1175 rows - see "head" and "glimpse" summarized herein.
-   I choose the dataset because I was thinking it would be interesting to identify the countries launching gadgets into space and the magnitude of the difference between the United States and other countries.
-   A graphical plot summary of the data is provided below - the plot presents an overview of the dataset and visually displays the number of launches per entity between 1957 and 2023.

```{r}
#| label: load-dataset
#| message: false

# cite: https://github.com/rfordatascience/tidytuesday/blob/master/data/2024/2024-04-23/readme.md

outer_space_objects <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-04-23/outer_space_objects.csv')

```

```{r}
dim(outer_space_objects)
```

```{r}
head(outer_space_objects)
```

```{r}
glimpse(outer_space_objects)
```

```{r}
filtered_outer_space_objects = outer_space_objects %>%
  filter (Entity %in% c("United States","China","Russia","Japan","France","India","European Space Agency", "Germany"))
```

## Graphical Summary of the Dataset

```{r}
ggplot(data = filtered_outer_space_objects, aes(x=Year, y=num_objects, color = Entity)) +
  geom_smooth(se = FALSE, method = loess) +
  labs(
    x = "year of launch",
    y = "number of objects launched",
    title = "Number of objects launched into space by year",
    subtitle = "data collected from United Nations Office of Outer Space Affairs",
    caption = "objects are defined as satelllites, probes, landers, crewed spacecrafts, and space station flights"
  ) +
  theme(
    axis.text.x = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8, face = "bold"),
    plot.title = element_text (size = 10, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 8, face = "plain", hjust = 0.5),
    plot.caption = element_text(size = 6, face = "plain", hjust = 1),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_line(color = "gray95")
  ) 
```

## Dataset Variable (column) Descriptions

-   The dataset comprises 4 columns and 1175 rows. The dataset summarizes space launch events between 1957 and 2023 as a function of the entity or country responsible for the launch.

-   **Entity**: <character> the country launching the space object

-   **Code**: <character> the type of gadget launched into Earth orbit or beyond (satellites, probes, landers, crewed spacecrafts, and space station components)

-   **Year**: <double-precision floating-point number> the year of the launch

-   **num_objects**: <double-precision floating-point number> the number of objects launched by "Entity" during a year.

## Questions

-   Question 1 - what country is most likely to become the first "space faring nation" and is there a close second? I propose to define "space faring nation" as the country that launches an order of magnitude more vehicles into space than all other countries combined. I'll fit a polynomial regression model to the dataset by country and extrapolate the model into the future to help identify the likely candidates.

## Analysis plan

-   The analysis plan is to plot the data on a map and display the launches by country over time. Currently, I have no idea how to do this, create a map, and plot an event from the data set onto the map?

-   I am thinking I will need to get a map of the world with the countries displayed. Another data set with the longitude and latitude for each country may be needed.

-   The launch event data set includes country and number of launches per year, to this I am thinking I will need to add columns defining the longitude and latitude of each launch event.

-   With longitude and latitude defined for each row/event I would then plot the launches on a map.

-   The analysis plan is to have an office hours meetign with Professor Chism the week of June 3rd and discuss how to plot country launch data on a world map. I'll then completed the required learning and apply that new skillset to the launch dataset.

## Detailed Analysis Plan

-   Review with Professor Chism how best to plot data on a world map

-   Use a polynomial regression model to extrapolate the launch data by country into the future for the six most prominant countries

-   Plot the extrapolated dataset onto a world map to identify the most likely country to become a "Space Faring Nation".

## Plan of Attack

| Task Name                                        | Status      | Date      | Summary                                                           |
|--------------------|-------------|-------------|----------------------------|
| Select dataset                                   | complete    | May 27th  | selected the Space Object Launch Dataset                          |
| Define question to be answered                   | complete    | May 28th  | what country is most likely to become the 1st space faring nation |
| Submit draft proposal for peer review            | complete    | May 29th  | proposal submitted - comments under review                        |
| Response to peer review                          | complete    | June 2nd  | peer review comments incorporate into the proposal                |
| Submit project proposal                          | complete    | June 3rd  | proposal submitted to Professor Chism                             |
| Office hours with Prof Chism to discuss approach | in-work     | June 9th  | need to learn how to plot launch data on a map?                   |
| Complete required learning                       | in-work     | June 15th | Reviewed material suggested by Prof. Chism                        |
| Complete initial analysis of the dataset         | not started | June 15th | Data plotted on a map                                             |
| draft presentation and writeup                   | not started | June 21st | draft files committed to github                                   |
| Complete presentation                            | not started | June 23rd | submit to github                                                  |
| Comlete writeup                                  | not started | June 23rd | submit to github                                                  |
| final submission                                 | not started | June 24th | submit to github                                                  |
|                                                  |             |           |                                                                   |
|                                                  |             |           |                                                                   |

: Approach by Task
