---
title: "Objects Launched into Space"
subtitle: "Proposal"
author: Derek Rice
  - name: "derekriceX"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "INFO526 Term Project"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
if (!require("pacman"))
    install.packages("pacman")

pacman::p_load(tidyverse, tidytuesdayR, dplyr, janitor, lubridate, numpy)

```

## Dataset: Outer Space Objects 

```{r}
#| label: load-dataset
#| message: false

# cite: https://github.com/rfordatascience/tidytuesday/blob/master/data/2024/2024-04-23/readme.md

# Option 1: tidytuesdayR package 
## install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2024-04-23')
## OR
tuesdata <- tidytuesdayR::tt_load(2024, week = 17)

outer_space_objects <- tuesdata$outer_space_objects

```

```{r}
head(outer_space_objects)
```

```{r}
glimpse(outer_space_objects)
```

```{r}

filtered_outer_space_objects = outer_space_objects %>%
  filter (Entity %in% c("United States","China","Russia","Japan","France","India","European Space Agency", "Germany"))
```

## **Description**

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

```{r}
ggplot(data = filtered_outer_space_objects, aes(x=Year, y=num_objects, color = Entity)) +
  geom_smooth(se = FALSE, method = loess) +
  labs(
    x = "year of launch",
    y = "number of objects launched",
    title = "Number of objects launched into space by year",
    subtitle = "data collected from United Nations Office of Outer Space Affairs",
    caption = "objects are defined as satelllites, probes, landers, crewed spacecrafts, and space station flights"
  ) +
  theme(
    axis.text.x = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8, face = "bold"),
    plot.title = element_text (size = 10, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 8, face = "plain", hjust = 0.5),
    plot.caption = element_text(size = 6, face = "plain", hjust = 1),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_line(color = "gray95")
  ) 
```

## Questions

The two questions you want to answer.

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
